---
title: "Plateau v2"
author: "Mark Brewer"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

## R code in support of the paper
## Brewer M.J., O'Hara R.B., Anderson B.J. & Ohlemüller R. (). Climate envelopes for species distribution models. Methods in Ecology and Evolution [In press].

This code reproduces the examples and figures from the paper, and hopefully helps in understanding the related R package "plateau", also available on GitHub, as the following code suggests:

```{r Install.plateau, echo=TRUE}
# Install the plateau package from GitHub - need devtools for this
library(devtools)
install_github("MarkJBrewer/plateau")
library(plateau)
```

## Figure 1

```{r Figure 1, echo=FALSE}
# Figure 1 - Example 1D envelope
op <- par()
par(mfrow = c(1,2), mar = c(3.2,3,0.5,1), mgp = c(2,1,0))
plot(0,0,xlim=c(-10,10),ylim=c(-10,10),type="n",
    xlab="",ylab="",
    yaxt="n",xaxt="n")
mtext("Linear Predictor",side=2,line=1,cex=1.1)
mtext("Climate Covariate",side=1,line=1,cex=1.1)
mtext(expression(beta[0]),side=2,line=0,cex=1.2)
lines(c(-10,-3),c(-10,0))
lines(c(-3,1),c(0,0))
lines(c(1,10),c(0,-10))
text(8,7,"(a)")
text(-8.4,-4,expression(paste(alpha[1],"+",beta[1],x)))
text(7.7,-4,expression(paste(alpha[2],"+",beta[2],x)))
text(1.5,2.7,expression(paste("(",a[x],",",a[z],")")))
abline(h=0,lty=3)
lines(c(-3,4),c(0,10),lty=2)
lines(c(1,-8),c(0,10),lty=2)
plot(0,0,xlim=c(-10,10),ylim=c(0,1),type="n",
    xlab="",ylab="",yaxt="n",xaxt="n")
mtext("Climate Covariate",side=1,line=1,cex=1.1)
mtext("Response (Probability)",side=2,line=1,cex=1.1)
lines(c(-3,1),0.8*c(1,1))
xplot <- seq(-10,-3,length=1000)
lines(xplot,0.8*21/(1+exp(-xplot)))
xplot <- seq(1,10,length=1000)
lines(xplot,0.8*3.7/(1+exp(xplot)))
text(8,0.85,"(b)")
par(op)
```

## Figure 2

```{r Figure 2, echo=FALSE}
# Figure 2 - Example bivariate envelope, at different stages of evolution
env.calc <- function(beta1.par,beta2.par,az,ax,beta0,gamma.par,N){
    x <- seq(-2,2,length=N)
    x.clim <- cbind(x,x)
    x.clim.C <- array(NA,dim=c(N,2))
    x.beta <- array(NA,dim=c(N,2,2))
    calc1 <- array(NA,dim=c(N,2))
    calc2 <- array(NA,dim=c(N,2))
    calcTotal.tmp <- array(NA,dim=c(N,N))
    calcTotal <- array(NA,dim=c(N,N))
    for(i in 1:N) {
        for(e in 1:2) {
            x.clim.C[i,e] <- x.clim[i,e]-ax[e]
            x.beta[i,e,1] <- -beta1.par[e]*x.clim.C[i,e]
            x.beta[i,e,2] <- beta2.par[e]*x.clim.C[i,e]
            calc1[i,e] <- max(x.beta[i,e,1],x.beta[i,e,2])
            calc2[i,e] <- calc1[i,e]*abs(x.clim.C[i,e])
        }
    }
    for(i1 in 1:N) {
        for(i2 in 1:N) {
            calcTotal.tmp[i1,i2] <- az - sqrt(calc2[i1,1] + calc2[i2,2] + gamma.par*x.clim.C[i1,1]*x.clim.C[i2,2])
            calcTotal[i1,i2] <- min(calcTotal.tmp[i1,i2],beta0)
        }
    }
    env <- list(calcTotal=calcTotal,x=x,maxcalc=max(calcTotal.tmp))
    return(env)
}
op <- par()
par(mfrow=c(5,2),mar=c(0.5,0.5,0.5,0.5))
N1 <- 20  # for persp
N2 <- 400 # for contour
# Cone development - standard cone
beta1.par <- c(10,10)
beta2.par <- c(10,10)
az <- 0
ax <- c(0,0)
beta0 <- 1
gamma.par <- 0
env <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N1)
persp(env$x,env$x,env$calcTotal, phi=30, theta=-40, border = 1,
    zlab="",xlab="",ylab="",axes=FALSE,d=2)
text(-0.5,-0.5,"(a)")
calcTotal <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N2)
contour(env$x,env$x,env$calcTotal-env$maxcalc,xaxt="n",yaxt="n",levels=seq(-2,-12,by=-2))
# Now stretch in one direction
beta1.par <- c(6,10)
beta2.par <- c(6,10)
env <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N1)
persp(env$x,env$x,env$calcTotal, phi=30, theta=-40, border = 1,
    zlab="",xlab="",ylab="",axes=FALSE,d=2)
text(-0.5,-0.5,"(b)")
calcTotal <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N2)
contour(env$x,env$x,env$calcTotal-env$maxcalc,xaxt="n",yaxt="n",levels=seq(-2,-12,by=-2))
# Now allow interaction terms along diagonal
beta1.par <- c(6,10)
beta2.par <- c(6,10)
gamma.par <- -5
env <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N1)
persp(env$x,env$x,env$calcTotal, phi=30, theta=-40, border = 1,
    zlab="",xlab="",ylab="",axes=FALSE,d=2)
text(-0.5,-0.5,"(c)")
calcTotal <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N2)
contour(env$x,env$x,env$calcTotal-env$maxcalc,xaxt="n",yaxt="n",levels=seq(-2,-12,by=-2))
# Now allow different slopes either side of apex
beta1.par <- c(5,10)
beta2.par <- c(9,40)
gamma.par <- -5
env <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N1)
persp(env$x,env$x,env$calcTotal, phi=30, theta=-40, border = 1,
    zlab="",xlab="",ylab="",axes=FALSE,d=2)
text(-0.5,-0.5,"(d)")
calcTotal <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N2)
contour(env$x,env$x,env$calcTotal-env$maxcalc,xaxt="n",yaxt="n",levels=seq(-2,-12,by=-2))
# Finally, allow the top-slice
beta1.par <- c(5,10)
beta2.par <- c(9,40)
gamma.par <- -5
beta0 <- -3
env <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N1)
persp(env$x,env$x,env$calcTotal, phi=30, theta=-40, border = 1,
    zlab="",xlab="",ylab="",axes=FALSE,d=2)
text(-0.5,-0.5,"(e)")
calcTotal <- env.calc(beta1.par,beta2.par,az,ax,beta0,gamma.par,N2)
contour(env$x,env$x,env$calcTotal-env$maxcalc,xaxt="n",yaxt="n",levels=seq(-2,-12,by=-2))
par(op)
```
## Figure 3

```{r Figure 3, echo=FALSE}
# Figure 3 - presence/absence plots of the two species studied in the paper
# Upper plot
par(mfrow=c(1,1))
species <- "fagsyl"
source("data_load.R")
species.name <- "Fagus sylvatica" # fagsyl
map.plot(inputs=AFEData[,species]==1,plot.type="Presence",coordinates=coords.list,
    species.name=species.name)
# Lower plot
par(mfrow=c(1,1))
species <- "quecoc"
source("data_load.R")
species.name <- "Quercus coccifera" # quecoc
map.plot(inputs=AFEData[,species]==1,plot.type="Presence",coordinates=coords.list,
    species.name=species.name)
```

## Figure 4
```{r Figure 4, echo=FALSE}
# Figure 4 - a four panel plot comparing GAM envelopes with our new envelopes
species <- "fagsyl"
source("data_load.R")
use.var <- c("tc","tw")
gamx1 <- AFEData[,use.var[1]]
gamx2 <- AFEData[,use.var[2]]
x1 <- seq(min(gamx1),max(gamx1),length=101)
x2 <- seq(min(gamx2),max(gamx2),length=101)
x1.plot <- rep(x1,101)
x2.plot <- rep(x2,each=101)
newgamdat <- data.frame(x1=x1.plot,x2=x2.plot)
gamy <- AFEData[,species]
gamdata <- data.frame(y=gamy,x1=gamx1,x2=gamx2)
# Bottom left plot
gamfit2 <- gam(y~te(x1,x2),family=binomial,data=gamdata)
predgam2 <- predict(gamfit2,newdata=newgamdat,type="response")
contour(x1,x2,matrix(predgam2,ncol=101),xlab=EnvLabels[use.var[1]],ylab=EnvLabels[use.var[2]])
points(AFEData[,use.var[1]],AFEData[,use.var[2]],col=as.numeric(AFEData[,species])+1,pch=16,cex=0.7)
title("Default k, using te()")
# Bottom right plot
gamfit3 <- gam(y~te(x1,x2,sp=c(0.01,0.01)),family=binomial,data=gamdata)
predgam3 <- predict(gamfit3,newdata=newgamdat,type="response")
contour(x1,x2,matrix(predgam3,ncol=101),xlab=EnvLabels[use.var[1]],ylab=EnvLabels[use.var[2]])
points(AFEData[,use.var[1]],AFEData[,use.var[2]],col=as.numeric(AFEData[,species])+1,pch=16,cex=0.7)
title("Default k, using te() with sp=0.01")
# Top plots - need to work out the envelopes using WinBUGS
initial.pars <- c(4,4,4,4,0.5,0.5,1,1,-4)
fagus.glm.1 <- fit.glm.env(y=AFEData[,species],x.clim=AFEData[,use.var],
    initial.pars.input=initial.pars)
initial.pars <- fagus.glm.1$par
# Note - the WinBUGS takes a few hours to run here
fagus.bugs.1 <- fit.bugs.env(y=AFEData[,species],x.clim=AFEData[,use.var],
    car.sigma=0.1,adj=adj,num=num,u=u,
    u.clique.start=u.clique.start,u.clique.end=u.clique.end,
    adj.clique.start=adj.clique.start,adj.clique.end=adj.clique.end,
    clique=in.clique,clique.i=in.clique.newnum,
    informative.priors=list(ax=TRUE),
    initial.pars.input=initial.pars,burnin=4000,post.burnin=1000,chains=2,thin=1,
    working.directory=NULL,WinBUGS.debug=FALSE,
    estimate.p=FALSE,estimate.u=FALSE)
# Top left plot
envelope.plot(fagus.bugs.1,type="persp",x.labels=EnvLabels[use.var],len=40)
# Top right plot
envelope.plot(fagus.bugs.1,type="contour",x.labels=EnvLabels[use.var])
```

## Figure 5
```{r Figure 5, echo=FALSE}
# Figure 5 - showing 2D envelopes for different values of a third variable
use.var <- c("dro","tc","tw")
initial.pars <- c(4,4,4,4,4,4,0.5,0.5,0.5,1,1,4,4,4)
fagus.glm.2 <- fit.glm.env(y=AFEData[,species],x.clim=AFEData[,use.var],
    initial.pars.input=initial.pars)
initial.pars <- fagus.glm.2$par
# Note - the WinBUGS takes a few hours to run here
fagus.bugs.2 <- fit.bugs.env(y=AFEData[,species],x.clim=AFEData[,use.var],
    car.sigma=0.1,adj=adj,num=num,u=u,
    u.clique.start=u.clique.start,u.clique.end=u.clique.end,
    adj.clique.start=adj.clique.start,adj.clique.end=adj.clique.end,
    clique=in.clique,clique.i=in.clique.newnum,
    constrain.beta=matrix(c(TRUE,FALSE,FALSE,FALSE,FALSE,FALSE),ncol=2), # read down columns
    initial.pars.input=initial.pars,burnin=4000,post.burnin=1000,chains=2,thin=1,
    working.directory=NULL,WinBUGS.debug=FALSE,
    estimate.p=FALSE,estimate.u=FALSE)
# Top left plot
envelope.plot(fagus.bugs.2,type="contour",x.labels=EnvLabels[use.var],plot.vars=c("dro","tw"),
    close.points="plot",fix.values=c(tc=-6))
title(expression(paste("Mean Temp of Coldest Month = ",-6,degree,"C")))
# Top right plot
envelope.plot(fagus.bugs.2,type="persp",x.labels=EnvLabels[use.var],plot.vars=c("dro","tw"),
    close.points="plot",fix.values=c(tc=-6),len=40)
title(expression(paste("\nMean Temp of Coldest Month = ",-6,degree,"C")),line=3)
# Middle left plot
envelope.plot(fagus.bugs.2,type="contour",x.labels=EnvLabels[use.var],plot.vars=c("dro","tw"),
    close.points="plot",fix.values=c(tc=-2))
title(expression(paste("Mean Temp of Coldest Month = ",-2,degree,"C")))
# Middle right plot
envelope.plot(fagus.bugs.2,type="persp",x.labels=EnvLabels[use.var],plot.vars=c("dro","tw"),
    close.points="plot",fix.values=c(tc=-2),len=40)
title(expression(paste("\nMean Temp of Coldest Month = ",-2,degree,"C")),line=3)
# Bottom left plot
envelope.plot(fagus.bugs.2,type="contour",x.labels=EnvLabels[use.var],plot.vars=c("dro","tw"),
    close.points="plot",fix.values=c(tc=6))
title(expression(paste("Mean Temp of Coldest Month = ",+6,degree,"C")))
# Bottom right plot
envelope.plot(fagus.bugs.2,type="persp",x.labels=EnvLabels[use.var],plot.vars=c("dro","tw"),
    close.points="plot",fix.values=c(tc=6),len=40)
title(expression(paste("\nMean Temp of Coldest Month = ",+6,degree,"C")),line=3)
```

## Figure 6

```{r Figure 6, echo=FALSE}
# Figure 6 - maps of climate projections, for the two different WinBUGS model fits
# Left plot
use.var <- c("tc","tw")
xc <- AFEData[,use.var]
map.plot(inputs=fagus.bugs.1$par,x.clim.new=xc,coordinates=coords.list,
    species.name=species.name,scenario.name="Current")
# Right plot
use.var <- c("dro","tc","tw")
xc <- AFEData[,use.var]
map.plot(inputs=fagus.bugs.2$par,x.clim.new=xc,coordinates=coords.list,
    species.name=species.name,scenario.name="Current")
```

## Figure 7

```{r Figure 7, echo=FALSE}
# Figure 7 - envelopes for the second species, with the bottom plots illustrating the use of informative priors
species <- "quecoc"
source("data_load.R")
use.var <- c("dro","tc")
species.name <- "Quercus coccifera" # quecoc
initial.pars <- c(4,4,4,4,0.5,0.5,1,1,-4)
quercus.glm.1 <- fit.glm.env(y=AFEData[,species],x.clim=AFEData[,use.var],
    initial.pars.input=initial.pars)
initial.pars <- quercus.glm.1$par
quercus.bugs.1 <- fit.bugs.env(y=AFEData[,species],x.clim=AFEData[,use.var],
    car.sigma=0.1,adj=adj,num=num,u=u,
    u.clique.start=u.clique.start,u.clique.end=u.clique.end,
    adj.clique.start=adj.clique.start,adj.clique.end=adj.clique.end,
    clique=in.clique,clique.i=in.clique.newnum,
    informative.priors=list(ax=TRUE),
    initial.pars.input=initial.pars,burnin=5000,post.burnin=1000,chains=2,thin=1,
    working.directory=NULL,WinBUGS.debug=FALSE,
    estimate.p=FALSE,estimate.u=FALSE)
# Top left plot
envelope.plot(quercus.bugs.1,type="contour",x.labels=EnvLabels[use.var],x.plot.lims=c(0,1.5))
# Top right plot
envelope.plot(quercus.bugs.1,type="persp",x.labels=EnvLabels[use.var],x.plot.lims=c(0,1.5),len=40)
# The informative prior analysis - needs to use user-supplied WinBUGS code
# Load GBIF data
gbif <- read.csv("plateau_finaldata.csv")
# Very rough map of locations
with(gbif,plot(long,lat,col=1+(quecoc_pa==1),cex=0.3,pch=16))
# Rough histogram of temperatures as locations
with(gbif,hist(bio11[quecoc_pa==1]/10))
# Now as kernel density
with(gbif,plot(density(bio11[quecoc_pa==1]/10)))
# Work out likely 2% threshold
gbif.fit <- density(gbif$bio11[gbif$quecoc_pa==1]/10)
gbif.sample <- rnorm(100000, sample(gbif$bio11[gbif$quecoc_pa==1]/10, size = 100000, replace = TRUE), gbif.fit$bw)
100*mean(gbif.sample>13) # Roughly 2%
MTCOmax <- max(AFEData$tc)
MTCOmin <- min(AFEData$tc)
# Do same standardisation as plateau code - mapping to [0,1]
(13-MTCOmin)/(MTCOmax-MTCOmin)
initial.pars <- quercus.glm.1$par
# Run supplied code file for informative prior
quercus.bugs.2 <- fit.bugs.env(y=AFEData[,species],x.clim=AFEData[,use.var],
    car.sigma=0.1,adj=adj,num=num,u=u,
    u.clique.start=u.clique.start,u.clique.end=u.clique.end,
    adj.clique.start=adj.clique.start,adj.clique.end=adj.clique.end,
    clique=in.clique,clique.i=in.clique.newnum,
    informative.priors=list(ax=TRUE),
    initial.pars.input=initial.pars,burnin=5000,post.burnin=1000,chains=2,thin=1,
    working.directory=NULL,WinBUGS.debug=FALSE,
    estimate.p=FALSE,estimate.u=FALSE,
    WinBUGS.code="WinBUGSinform.txt",
    no.starting.value=list("beta[2,2]"))
# Bottom left plot
envelope.plot(quercus.bugs.2,type="contour",x.labels=EnvLabels[use.var],x.plot.lims=c(0,1.5))
# Bottom right plot
envelope.plot(quercus.bugs.2,type="persp",x.labels=EnvLabels[use.var],x.plot.lims=c(0,1.5),len=40)
```
